\section{Description of array semantics}
\label{ch:arr-semantics-appx}

\subsection{Value type}

Arrays offer semantic differences, that may make it more appealing to
build data structures from for certain use cases. We've seen that arrays are
value types, and as such, copies can be easily made through simple assignment to
a variable or passing into a function. While cloning a single slice is not much
more difficult, if we have a large, nested data structure, that is slice-based,
performing a deep-copy is a verbose and manual process (there is no standard
library deep-copy function as of Go 1.21). If instead, the data structure was
array-based, with no internal pointers (only potentially a top-level pointer to
the data structure), performing a deep copy becomes trivial.

\subsection{Comparison}

Arrays are comparable, while slices are not. Again, for a simple example, one
could define a function that compares the size and elements of slices to
determine if two slices are equal in value (not reference). It is slightly more
difficult to do the same with a nested slice-based data structure (although this
time around, the standard library does provide \texttt{reflect.DeepEqual} with
some caveats\footnote{\href{https://pkg.go.dev/reflect\#DeepEqual}{\texttt{reflect.DeepEqual}} differentiates between a nil
    and an empty slice, despite them semantically being equivalent.
    \href{https://pkg.go.dev/github.com/google/go-cmp/cmp\#Equal}{\texttt{cmp.Equal}}
    with the
    \href{https://pkg.go.dev/github.com/google/go-cmp/cmp/cmpopts\#EquateEmpty}{\texttt{cmpopts.EquateEmpty}}
    option can be used instead to equate semantically empty slices
    \autocite{deepEqual}.}), a nested array-based data structure (with no internal
pointers) can be simply compared with the \texttt{==} operator.

This comparable attribute of arrays becomes extremely important when we wish to
use a collection of elements as a key in a hash map. Only comparable data
structures can be used as map keys, which means that arrays can be used as map
keys, while slices cannot. This also applies recursively, so slice-based data
structures cannot be used as map keys, while array-based ones can.
